Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-02 submission/manuscript.Rmd
update EpiLC text and figure

We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes with knockout of _Kdm5c_ (Supplementary table 1).

@@

Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by assessing their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].

We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B). We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).

Using our curated list, we found `r format(nrow(epigerm))` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D). To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E). We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). Contrastingly, the _Kdm5c_-KO hippocampus instead expressed genes involved in late-stage sperm development, those involved in the sperm axoneme. This suggests the germline developmental program may be occuring ectopically as _Kdm5c_-KO cells progresses through somatic tissue development **(note: placement is awkward. Even put here or talk about in dicussion because ChIPseq could help support this? - KDM5C not bound to a lot of brain DEGs)**.

While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo. However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs does not revert cells back to a 2-cell state (Figure 3G).

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_-KO cellular function. We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including key drivers of germline identity that can be translated into protein.

>>>>

We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes with loss of _Kdm5c_ (Supplementary table 1).

@@

Misexpression of germline genes in the adult _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs through DESeq2 analysis of our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Using our curated list, we found _Kdm5c_-KO EpiLCs aberrantly expressed `r format(nrow(epigerm))`germline-enriched genes, including the previously characterized hippocampal DEG[@scandagliaLossKdm5cCauses2017] _Cytochrome C, testis-specific (Cyct)_ (Figure 3C). Although we observed aberrant expression of many tissue-enriched genes, we did not observe any significant difference in primed pluripotency genes (Figure 3D) or gross changes in _Kdm5c_-KO cell morpholgy during differentiation (Figure 3E), indicating KDM5C loss does not impair EpiLC formation.

To evaluate if all germline DEGs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across all sequencing datasets (Figure 3F). We then compared the function of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus showed strong enrichment of late-stage sperm genes, such those involved in the sperm axoneme. This shift from meiotic genes to later spermatogenesis genes in the hippocampus suggests the germline developmental program could occur ectopically as _Kdm5c_-KO cells progresses through somatic tissue development. **note: this is strengthened by the ChIP-seq data since KDM5C is not directly bound to many brain/flagellar DEGs.This point might be stronger in the ChIPseq figure**

Interestingly, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo. However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function. We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3H). We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
